Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-11-26 inst/paper/main.tex

Most dynamic languages provide their users with a facility to
transform unstructured text into executable code, and to evaluate that
code. In this paper we refer to this reflective facility of a language
as \eval bowing to its origins in LISP in 1956~\cite{lisp}. \Eval has
been much maligned over the years. In computing lore, it is as close
to a boogeyman as it gets. Yet, for McCarthy, \eval was simply the way
to write down the definition of LISP, he was surprised that someone
coded it up and offered it end users.  Since then, reflective
facilities have been used to parameterize programs over code patterns
that can be provided after the program is written.  The presence of
such a feature in a language is a hallmark of dynamism as it is a form
of delayed binding as what any particular call to \eval will do is
only know at program run-time when that particular call site is
evaluated.

@@

\vspace{2mm}\noindent\emph{The here and now.}  In this study, we
investigate the usage of \eval in programs written in the R
programming language. R is language designed by statisticians for
applications in data science~\cite{r}. What makes looking at R after
JavaScript interesting is that, while both languages are dynamic, they
are quite different. While one can program in an object-oriented style
in R, like in JavaScript, R is primarily a lazy, untyped, functional
language.  JavaScript was designed to run untrusted code in browser,
while R is used for statistical computing on desktops. JavaScript is a
general purpose language used by a wide community of programmers;
while R is used for scientific computing by data scientists and domain
experts with, often, limited programming experience. One can
distinguish between library implementers, developers with some
programming experience and a working knowledge of R, and end-users,
who are typically not expert programmers and often have only a cursory
knowledge of the language.\footnote{Consider that R evaluates function
argument lazily, just like Haskell. We informally surveyed end-users,
including computer scientists, and did not find a single user aware of
this fact. Library developers, on the other hand, know about laziness
and program defensively around it.} Our goal is thus to highlight the
differences in usage between JavaScript and R, and try to explain
those differences in terms of language features, application domain
and programmer experience.

@@

\section{Background and Related Work}

The origins of \eval go back to the first implementation of the Lisp
programming language~\cite{lisp}. Over the years, \eval has been included in
many dynamic languages, with some variability in expressive power and
nomenclature. In all of its incarnation \eval allows developers to
dynamically extend the code of their program at run-time.  Originally, \eval
was limited to interpreted languages, but with the advent of just-in-time
compilation this restriction was gradually lifted.

The intuition behind the soundiness manifesto~\cite{soundy} is that while
the use of a single \eval may invalidate every single invariant computed by
a static analyzer, this worst case usually does not happen.  In many cases,
\eval is used carefully to limited effect. This hypothesis must be validated
and, for it to be useful, we need to have a way to make some educated
guesses as to which side-effects will result from any particular \eval.

\section{R and The Design of Eval}

The R Project is a tool for data analysis.  At its heart it is a {vectorized,
  dynamic, lazy, functional, object-oriented} programming language with an
unusual combination of features~\cite{ecoop12} designed to ease learning by
non-programmers and enable rapid development of statistical
methods~\cite{R96}.  The language was designed as a successor to
S~\cite{S88}.

Most R data types are vectorized. Values are constructed by the \c{c(...)}
function, e.g., \c{c("hi","ho")} creates a vector of two strings.  R does
not differentiate scalars from vectors. To enable equational reasoning, R
copies values accessible through multiple variables when they are updated.
Values can be tagged by user-defined attributes. For instance, one can
attach the attribute \c{dim} to the value \c{x<-c(1,2,3,4)} with
\c{attr(x,"dim")<-c(2,2)}.  This causes arithmetic functions to treat \c x
as a $2 \times 2$ matrix. Another attribute is \c{class} which can be bound to a list
of names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
\c{human}; classes are used for object-oriented dispatch.  Every R
linguistic construct is desugared to a function call, even control flow
statements, assignments, and bracketing. Furthermore, all functions can be
redefined in user code. This makes R both flexible and challenging to
compile~\cite{dls19}.

Arguments to user-defined functions are bundled into thunks called
promises. Logically, a promise combines an expression's code, its
environment, and its value.  To access the value of a promise, one must
force it. Forcing a promise triggers evaluation and the computed value is
captured for future reference.

@@

\begin{figure*}[!t]\centering\includegraphics[width=.8\linewidth]
{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
\end{figure*}

@@

@@

\begin{figure}[!h]\centering\includegraphics[width=\linewidth]
  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
\end{figure}

>>>>

Most dynamic languages provide their users with a facility to
transform unstructured text into executable code, and to evaluate that
code. In this paper we refer to this reflective facility of a language
as \eval bowing to its origins in LISP in 1956~\cite{lisp}. \Eval has
been much maligned over the years. In computing lore, it is as close
to a boogeyman as it gets. Yet, for McCarthy, \eval was simply the way
to write down the definition of LISP, he was surprised that someone
coded it up and offered it to end users.  Since then, reflective
facilities have been used to parameterize programs over code patterns
that can be provided after the program is written.  The presence of
such a feature in a language is a hallmark of dynamism as it is a form
of delayed binding as what any particular call to \eval will do is
only know at program run-time when that particular call site is
evaluated.

@@

\vspace{2mm}\noindent\emph{The here and now.}  In this study, we
investigate the usage of \eval in programs written in the R
programming language. R is language designed by statisticians for
applications in data science~\cite{r,R96}. What makes looking at R after
JavaScript interesting is that, while both languages are dynamic, they
are quite different. While one can program in an object-oriented style
in R, like in JavaScript, R is primarily a lazy, untyped, functional
language.  JavaScript was designed to run untrusted code in browser,
while R is used for statistical computing on desktops. JavaScript is a
general purpose language used by a wide community of programmers;
while R is used for scientific computing by data scientists and domain
experts with, often, limited programming experience. One can
distinguish between library implementers, developers with some
programming experience and a working knowledge of R, and end-users,
who are typically not expert programmers and often have only a cursory
knowledge of the language.\footnote{Consider that R evaluates function
argument lazily, just like Haskell. We informally surveyed end-users,
including computer scientists, and did not find a single user aware of
this fact. Library developers, on the other hand, know about laziness
and program defensively around it.} Our goal is thus to highlight the
differences in usage between JavaScript and R, and try to explain
those differences in terms of language features, application domain
and programmer experience.

@@

\section{Background and Previous work}

\subsection{R in a nutshell}

R is a lazy functional programming language~\cite{ecoop12} with
dynamic features that allow to write object-oriented code.  Most data
types are vectorized. Values are constructed by the \c{c(...)}
function, e.g. \c{c("hi","ho")} creates a vector of two strings.  To
enable equational reasoning, R copies values accessible through
multiple variables when they are written to.  Values can be tagged by
user-defined attributes. For instance, one can attach the attribute
\c{dim} to the value \c{x<-c(1,2,3,4)} with \c{attr(x,"dim")<-c(2,2)}.
This causes arithmetic functions to treat \c x as a $2 \times 2$
matrix. Another attribute is \c{class} which can be bound to a list of
names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
\c{human}; classes are used for object-oriented dispatch.  Every R
linguistic construct is desugared to a function call, even control
flow statements, assignments, and bracketing. Furthermore, all
functions can be redefined in user code. This makes R both flexible
and challenging to compile~\cite{dls19}.  Arguments to user-defined
functions are bundled into thunks called \emph{promises}. Logically, a
promise combines an expression's code, its environment, and its value.
To access the value of a promise, one must force it. Forcing a promise
triggers evaluation and the computed value is captured for future
reference.

\subsection{Eval Semantics}

@@

@@

\subsection{Previous Work}
Originally, \eval was limited to interpreted languages, but with the
advent of just-in-time compilation this restriction was gradually
lifted.

\begin{figure}[!t]\hspace{-5mm}\includegraphics[width=1.05\linewidth]
{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
\end{figure}

@@

\begin{figure}[!h]\centering\includegraphics[width=.7\linewidth]
  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
\end{figure}
